---
title: 'EDA demo'
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(tidyverse)
library(rstan)
library(mnormt)
library(plgp)
library(gstat)
library(sp)
library(datasets)
library(SemiPar)
library(ggmap)
library(loo)
set.seed(02282021)
options(mc.cores = parallel::detectCores())
```

# Scallops Data Example

1. Create a test and training set from the scallop dataset. Then create a visual to show the test / training datasets.


```{r, message = F}
data(scallop)
scallop <- scallop %>% mutate(log.catch = log(tot.catch + 1), 
                              id = 1:n())

data_fig <- scallop %>% ggplot(aes(x=longitude, y = latitude, color = log.catch)) + geom_point() +  theme_bw() + scale_color_gradientn(colours = colorspace::diverge_hcl(7))
data_fig

scallop_train <- scallop %>% sample_n(100)
scallop_test <- scallop %>% filter(!id %in% scallop_train$id)
```


```{r}
data_fig + geom_point(aes(y = latitude, x = longitude), inherit.aes = F, data = scallop_test, color = "black", size = 3)+ geom_point(aes(y = latitude, x = longitude), color = "white", inherit.aes = F, data = scallop_test, size = 2) + geom_point()
```

2. Write code to estimate a non-spatial model using just the mean structure. Then construct a figure that includes mean predictions for each site in the test dataset.

```{r}
x <- cbind(scallop_train$latitude,scallop_train$longitude)
d <- sqrt(plgp::distance(x))
num_preds <- nrow(scallop_test)
x_preds <- cbind(scallop_test$latitude,scallop_test$longitude)
d_preds <- sqrt(plgp::distance(x_preds))
d_12 <- sqrt(plgp::distance(x, x_preds))
```

```{stan output.var='mean_model', eval = F}
data {
  int<lower=0> N; // number of observed data points 
  vector[N] y; // observed response
  int<lower=0> N_preds; // number of predictive points
}

parameters {
  real<lower = 0> tausq;
  real mu;
}

transformed parameters{
  vector[N] mu_vec;
  vector[N] tausq_vec;
  for(i in 1:N) mu_vec[i] = mu;
  for(i in 1:N) tausq_vec[i] = tausq;
}

model {
  y ~ multi_normal(mu_vec ,diag_matrix(tausq_vec));
  mu ~ normal(0, 10);
  
}

generated quantities {
  vector[N_preds] y_preds;
  vector[N_preds] mu_preds;
  vector[N_preds] tausq_preds;

  for(i in 1:N_preds) mu_preds[i] = mu;
  for(i in 1:N_preds) tausq_preds[i] = tausq;

  y_preds = multi_normal_rng(mu_preds, diag_matrix(tausq_preds));
}

```



```{r, results = 'hide'}
mean_surface <- stan("mean_regression.stan", 
                  data=list(N = nrow(scallop_train), 
                            y = scallop_train$log.catch,
                            N_preds = num_preds),
                  iter = 5000)
```

```{r}
print(mean_surface, pars = c('mu', 'tausq', 'y_preds[1]','y_preds[2]','y_preds[3]'))

mean_preds <- colMeans(extract(mean_surface)['y_preds']$y_preds)

pred_df <- scallop_test %>% bind_cols(tibble(preds = mean_preds))

data_fig + geom_point(aes(y = latitude, x = longitude), color = "black", inherit.aes = F, data = pred_df, size = 3) + geom_point(aes(y = latitude, x = longitude), color = "white", inherit.aes = F, data = pred_df, size = 2) +
geom_point(aes(y = latitude, x = longitude, color = preds), data = pred_df, size = 1, inherit.aes = F) 
```

3. Now fit a model with spatial structure and construct a figure that includes mean predictions for each site in the test dataset.

```{stan output.var="spatial_regression_chol", eval = F}
data {
  int<lower=0> N; // number of observed data points 
  vector[N] y; // observed response
  matrix[N,N] dist; // observed distance matrix
  real phi_lower; // lower point for phi (range)
  real phi_upper; // upper point for phi (range)
  int<lower=0> N_preds; // number of predictive points
  matrix[N_preds,N_preds] dist_preds; // distance matrix for predictive points
  matrix[N, N_preds] dist_12; //distance between observed and predicted
  real phi_a;
  real phi_b;
  real sigmasq_a;
  real sigmasq_b;
  real tausq_a;
  real tausq_b;
}

parameters {
  real<lower = phi_lower, upper = phi_upper> phi;
  real<lower = 0> sigmasq;
  real<lower = 0> tausq;
  real mu;
}

transformed parameters{
  vector[N] mu_vec;
  vector[N] tausq_vec;
  corr_matrix[N] Sigma;
  
  for(i in 1:N) mu_vec[i] = mu;
  for(i in 1:(N-1)){
   for(j in (i+1):N){
     Sigma[i,j] = exp((-1)*dist[i,j]/ phi);
     Sigma[j,i] = Sigma[i,j];
   }
 }
 for(i in 1:N) Sigma[i,i] = 1;
 for(i in 1:N) tausq_vec[i] = tausq;
}

model {
  matrix[N, N] L;
  L = cholesky_decompose(sigmasq * Sigma + diag_matrix(tausq_vec));

  y ~ multi_normal_cholesky(mu_vec, L);
  phi ~ inv_gamma(phi_a, phi_b);
  sigmasq ~ inv_gamma(sigmasq_a, sigmasq_b);
  tausq ~ inv_gamma(tausq_a, tausq_b);
  mu ~ normal(0, 10);
}

generated quantities {
  vector[N_preds] y_preds;
  vector[N] y_diff;
  vector[N_preds] mu_preds;
  corr_matrix[N_preds] Sigma_preds;
  vector[N_preds] tausq_preds;
  matrix[N, N_preds] Sigma_12;

  for(i in 1:N_preds) tausq_preds[i] = tausq;
  for(i in 1:N_preds) mu_preds[i] = mu;
  for(i in 1:N) y_diff[i] = y[i] - mu;
  

  for(i in 1:(N_preds-1)){
   for(j in (i+1):N_preds){
     Sigma_preds[i,j] = exp((-1)*dist_preds[i,j]/ phi);
     Sigma_preds[j,i] = Sigma_preds[i,j];
   }
 }
 for(i in 1:N_preds) Sigma_preds[i,i] = 1;
 
   for(i in 1:(N)){
   for(j in (1):N_preds){
     Sigma_12[i,j] = exp((-1)*dist_12[i,j]/ phi);
   }
 }

 y_preds = multi_normal_rng(mu_preds + (sigmasq * Sigma_12)' * inverse(sigmasq * Sigma) * (y_diff), sigmasq * Sigma_preds + diag_matrix(tausq_preds) - (sigmasq * Sigma_12)' * inverse(sigmasq * Sigma + diag_matrix(tausq_vec)) * (sigmasq * Sigma_12) );
}

```


```{r, results = 'hide'}
spatial_surface <- stan("spatial_regression_chol.stan", 
                  data=list(N = nrow(scallop_train), 
                            y = scallop_train$log.catch,
                            dist = d,
                            phi_lower= .05,
                            phi_upper = 2.5,
                            N_preds = num_preds,
                            dist_preds = d_preds,
                            dist_12 = d_12,
                            phi_a = 1,
                            phi_b = 1,
                            sigmasq_a = 3,
                            sigmasq_b = 3,
                            tausq_a = 3,
                            tausq_b = 3),
                 chains = 2)
```


```{r}
print(spatial_surface, pars = c('mu', 'sigmasq','tausq', 'y_preds[1]','y_preds[2]','y_preds[3]'))


spatial_preds <- colMeans(extract(spatial_surface)['y_preds']$y_preds)

pred_df_spatial <-  pred_df %>% bind_cols(tibble(spatial_preds = spatial_preds))

data_fig + geom_point(aes(y = latitude, x = longitude), color = "black", inherit.aes = F, data = pred_df, size = 3) + geom_point(aes(y = latitude, x = longitude), color = "white", inherit.aes = F, data = pred_df_spatial, size = 2) +
geom_point(aes(y = latitude, x = longitude, color = spatial_preds), data = pred_df_spatial, size = 1, inherit.aes = F) 
```


4. Compare the predictive ability of the spatial and non-spatial model

```{r}
pred_df_spatial %>% mutate(diff_mean = tot.catch - exp(preds),
                           diff_spatial = tot.catch - exp(spatial_preds)) %>%
  summarise(rmse_mean = sqrt(mean(diff_mean^2) ),
            rmse_spatial = sqrt(mean(diff_spatial^2)),
            mad_mean = mean(abs(diff_mean)),
            mad_spatial = mean(abs(diff_spatial)))
```


### Other model fitting options

- `krige()` in the `gstat` package. However, the spatial structure needs to be extracted from fitting a variogram to the empirical variogram.

- `geoR` has `krige.conv()` and `krige.bayes()` options see textbook for more.

- `spBayes` is another package with quite a few options

