---
title: 'EDA demo'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(tidyverse)
library(rstan)
library(mnormt)
library(plgp)
library(gstat)
library(sp)
library(datasets)
library(SemiPar)
library(ggmap)
library(loo)
set.seed(02282021)
options(mc.cores = parallel::detectCores())
```

# Scallops Data Example

1. Create a test and training set from the scallop dataset. Then create a visual to show the test / training datasets.


```{r, message = F}
data(scallop)
scallop <- scallop %>% mutate(log.catch = log(tot.catch + 1), 
                              id = 1:n())
```

2. Write code to estimate a non-spatial model using just the mean structure. Then construct a figure that includes mean predictions for each site in the test dataset.

3. Now fit a model with spatial structure and construct a figure that includes mean predictions for each site in the test dataset.

4. Compare the predictive ability of the spatial and non-spatial model


### Other model fitting options

- `krige()` in the `gstat` package. However, the spatial structure needs to be extracted from fitting a variogram to the empirical variogram.

- `geoR` has `krige.conv()` and `krige.bayes()` options see textbook for more.

- `spBayes` is another package with quite a few options

