---
title: "Spatial EDA"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F)
library(knitr)
library(ggplot2)
library(dplyr)
library(gstat)
library(sp)
library(datasets)
library(SemiPar)
library(ggmap)
set.seed(02192021)
```
# Exploratory Data Analysis

## EDA Overview
- Exploratory Data Analysis (EDA) is commonly used to explore and visualize data sets.
- EDA is not a formal analysis, but can inform modeling decisions.
- What are we interested in learning about with spatial data?

## Data Decomposition: Time Series
- In time series analysis, the first step in the EDA process was to decompose the observed data into a trend, seasonal cycles, and a random component.

```{r}
plot(decompose(AirPassengers))
```

\newpage

## Data Decomposition: Spatial Data
- Similarly spatial data will be decomposed into the mean surface and the error surface.

- For example, elevation and distance from major bodies of water would be part of the mean surface for temperature.

- The mean surface is focused on the global, or first-order, behavior.

- The error surface captures local fluctuations, or second-order, behavior.

## Response Surface vs. Spatial Surface
- Spatial structure in the response surface and spatial structure in the error surface are not one-and-the-same.

- $E[(Y(\boldsymbol{s})-\mu)(Y(\boldsymbol{s'})-\mu)]$ vs. $E[(Y(\boldsymbol{s})-\mu(\boldsymbol{s}))(Y(\boldsymbol{s'})-\mu(\boldsymbol{s}))]$

- There are stationarity implications for considering the residual surface.

- Data sets contain two general types of useful information: spatial coordinates and covariates.

- Regression models will be used to build the mean surface.

## Spatial EDA Overview

1. Map of locations
2. Histrogram or other distributional figure
3. 3D scatterplot
4. General Regression EDA
5. Variograms and variogram clouds
6. Anistopic diagnostics

\newpage

# Scallops Data Example

## 1. Map of Locations
```{r, message = F}
data(scallop)
scallop <- scallop %>% mutate(log.catch = log(tot.catch + 1))

scallops_box <- c(left = min(scallop$longitude)-.25, bottom = min(scallop$latitude)-.25, right = max(scallop$longitude)+.25, top =max(scallop$latitude)+.25)
scallops_map <- get_map(scallops_box) %>% ggmap()
  
scallops_map + geom_point(aes(x = longitude, y = latitude, color= tot.catch), data = scallop, inherit.aes = F)
```

## 1. Map of Locations - Takeaways

_Goal_: Understand the sampling approach

- Is this a grid?
- Are there directions that have larger distances?
- How large is the spatial extent?

\newpage

## 2. Histogram
```{r}
scallop %>% ggplot(aes(x=tot.catch)) + geom_histogram(bins=70) + ggtitle('histogram of scallop catch') + xlab('Number of Scallops') + theme_bw()
```

```{r}
scallop %>% ggplot(aes(x=log.catch)) + geom_histogram(bins=70) + ggtitle('histogram of log scallop catch (+1)') + xlab('Number of Scallops') + theme_bw()
```

## 2. Histogram - Takeaways

_Goal:_ Identify a sampling distribution for the data

- Continuous or discrete data
- A linear model approach will be used for the response
- Spatial structure can also be included in generalized linear models
- Outliers are worth investigating, but a data point that does not fit the assumed model should not automatically be eliminated

\newpage

## 3. 3D scatterplot
```{r}
scallop %>%  ggplot(aes(x=longitude, y = latitude)) + geom_point(aes(color=tot.catch)) + scale_colour_gradient2() + theme_bw()
```

```{r}
scallop %>% ggplot(aes(x=longitude, y = latitude)) + geom_point(aes(color=log.catch)) + scale_colour_gradient2() + theme_bw()
```

## 3. 3D scatterplot - Takeaways

_Goal:_ Examine the spatial pattern of the response

- Again, this is the response not the residual
- Can also think about a contour plot (using some interpolation method)

\newpage

## 4. General Regression EDA

- Assessing relationship between variable of interest and covariate information
- No covariates are present in the scallops data


## 5. Variograms and variogram clouds
## 5. Variograms 
```{r,  eval = T}
coordinates(scallop) = ~longitude+latitude
scallop.sp <- scallop
proj4string(scallop.sp) <- CRS("+proj=longlat +datum=WGS84")  ## for example
scallop.utm <- spTransform(scallop.sp, CRS("+proj=utm +zone=18 ellps=WGS84"))

plot(variogram(log.catch~1, scallop.utm))
```

## 5. Variogram Cloud
```{r}
plot(variogram(log.catch~1, scallop.utm, cloud=T))
```

## 5. Variograms and variogram clouds: Takeaways

_Goal:_ Visually diagnose spatial structure

\newpage 

## 6. Anisotropy

_Goal:_ Determine if direction influences spatial structure


## Directional Variogram
- All of the variograms we have looked at are isotropic

```{r}
plot(variogram(log.catch~1, scallop.sp, alpha = c(0, 45, 90, 135)))
```

## Separable Correlations Functions

- If the differences in spatial structure are directly related to two coordinate sets, we can create a stationary, anistropic covariance function

- Let $$cor(Y(\boldsymbol{s + h}), Y(\boldsymbol{s})) = \rho_1(h_y) \rho_2(h_x),$$
where $\rho_1()$ and $\rho_2()$ are proper correlation functions.
- A scaling factor, $\sigma^2,$ can be used to create covariance.

## Geometric Anistropy

- Another solution is the class of geometric anisotropic covariance functions with 
$$C(\boldsymbol{s} - \boldsymbol{s'}) = \sigma^2 \rho((\boldsymbol{s} - \boldsymbol{s'})^TB(\boldsymbol{s} - \boldsymbol{s'})),$$
where $B$ is positive definite matrix and $\rho$ is a valid correlation function
- $B$ is often referred to as a transformation matrix which rotates and scales the coordinates, such that the resulting transformation can be simplified to a distance.

## Sill, Nugget, and Range Anisotropy
- Recall the sill is defined as $\lim_{d \rightarrow \infty} \gamma(d)$
- Let $\boldsymbol{h}$ be an arbitrary separation vector, that can be normalized as $\frac{\boldsymbol{h}}{||\boldsymbol{h}||}$
- If $\lim_{a \rightarrow \infty} \gamma(a \times \frac{\boldsymbol{h}}{||\boldsymbol{h}||})$ depends on $\boldsymbol{h}$, this is referred to as sill anisotropy.
- Similarly the nugget and range can depend on $\boldsymbol{h}$ and give nugget anisotropy and range anisotropy 

